# Data wrangling with tidyverse and pandas {#wrangling}

```{r, echo = FALSE}
library(tidyverse)
library(palmerpenguins)
```

We've learned some strategies to subset, reshape and update data using base tools and functions. In this chapter, we'll use functions from R packages and Python libraries to wrangle our data in alternative ways that may be more efficient, intuitive and/or readable for collaborators.

Throughout this chapter, we'll see examples using the penguins data set in the palmerpenguins R package (make sure you have it installed and attached to follow along). 

## Meet the tidyverse and pandas

Names for things that seem weird now will be come so ingrained in your brain that you won't even realize the nonsense you're spewing to non-R or -Python users (we'll practice avoiding that throughout MEDS, but be warned...). A couple of big ones are **tidyverse** and **pandas** - what are they, and how are they useful for us? 

### R wrangling: `dplyr` & `tidyr` packages in the tidyverse

The [tidyverse](https://www.tidyverse.org/) is a collection of R packages built to make data wrangling and visualizations easier for everyone. The packages are *opinionated* - meaning that they use similar syntax so that functions from different packages within the `tidyverse` play nicely together - especially useful once you get into burlier wrangling. 

The two tidyverse packages that we'll focus on here are [dplyr](https://dplyr.tidyverse.org/) and [tidyr](https://tidyr.tidyverse.org/). The `dplyr` package contains powerful and intuitive functions for wrangling data (subsetting, filtering, choosing or moving columns, finding summary statistics by group, etc.). The `tidyr` package contains functions useful for reshaping data into tidy structure (e.g. converting from long-to-wide or wide-to-long format so that each variable is a single column, each row is a single observation, and each cell contains a single value). 

To get `dplyr` and `tidyr`, along with all other packages in the tidyverse, install the `tidyverse` by running `install.packages("tidyverse")` in the R Console. 

### Python wrangling: the `pandas` library 

The [pandas](https://pandas.pydata.org/) library is a "fast, powerful, flexible and easy to use open source data analysis and manipulation tool,
built on top of the Python programming language" (from https://pandas.pydata.org/). With tools for reading, wrangling, subsetting, and reshaping data, it is the go-to for data manipulation in Python. 

If you already have Anaconda installed, you're all set! The `pandas` library comes along with it (in addition to a bunch of other libraries useful for data wrangling and analysis, like `NumPy` and `matplotlib`). 

## Selecting columns

Make a subset of *columns* (variables, if data are tidy) using: 

- tidyverse: `dplyr::select()`
- pandas: 

### `dplyr::select()`

Use `dplyr::select()` in R to choose columns by numbered position or column name (recall: use `names()` to return a vector of all column names, or `View()` to bring up the data frame in a new viewing tab).

**Example:** Starting with the penguins data frame, create and store a subset that only contains variables `species` and `island`.

```{r}
penguin_sp_isl <- penguins %>% 
  select(species, island)

# Alternative (but I like the formatting above...): 
penguin_sp_isl <- select(penguins, species, island)
```

Which return a data frame only containing those two variables. Let's check using `names()`:

```{r}
names(penguin_sp_isl)
```

### Selecting columns in Python

We choose columns in Python using the syntax `my_df[['col_a', 'col_b']]`, which will select `col_a` and `col_b` from a pandas DataFrame stored as `my_df`. 

In the Python example below, we store the `penguins` data frame as a Python DataFrame, then create a subset that only contains `species` and `island`: 

```{python}
# Store the existing R object `penguins` as a Python DataFrame:
py_penguins = r.penguins

# Then make a subset only containing `species` and `island`:
penguin_sp_isl = py_penguins[["species", "island"]]

# Check out the first few lines to see variables included:
penguin_sp_isl.head()
```
## Subsetting rows based on conditions

Creating subsets that match conditions you specify will absolutely be one of the most common data wrangling things you do. 

### `dplyr::filter()`

Create subsets of data based on...

### Subsetting rows in Python


## Adding columns

### `dplyr::mutate()`


## Finding grouped statistics


## Joining data
