# Data visualization with `ggplot2` and `seaborn`

```{r, include = FALSE}
library(tidyverse)
library(palmerpenguins)
```

GET SEABORN WORKING
```{python}
# import seaborn as sns
```

Visualizing data is a critical skill in every step of data science - from exploratory visualization to look for major patterns, groups or outliers in the data to preparing highly designed infographics for stakeholders, every data scientist must have technical and conceptual skills in data viz. 

In this chapter, we'll learn about two powerful packages for data visualization: 

- `ggplot2`: the go-to data visualization package in R (part of the tidyverse)
- `seaborn`: one of several popular data viz libraries in Python

## The grammar of graphics

The "gg" in `ggplot2` is for the **grammar of graphics** - a way to describe parts of a graph [@layered-grammar], which in turn provides us an organized and predictable plan to build data visualizations. Understanding the grammar of graphics, and some vocabulary, can help to demystify how we build visualizations with code and the role of some specific functions. 

## Basic anatomy of graphs with `ggplot2` in R, and `seaborn` in Python

### `ggplot2` bare minimum 

Making the most basic ggplot requires 3 pieces:

1. Tell R you're making a ggplot graph (`ggplot()`)
2. Tell it what data you're plotting (including what the x- and/or y-axis variables are)
3. Tell it what type of plot you want to make

**Note:** it is very important to carefully consider the type of data you're trying to visualize when making a graph. Some visualizations are not appropriate for some types of data. We'll discuss this a bit more throughout the class, but one of my favorite resources to consider appropriate visualizations based on data type is [From Data to Viz](data-to-viz.com).

For example, let's make a basic scatterplot of bill length versus bill depth from the `penguins` dataset (in the `palmerpenguins` package).  

```{r}
ggplot(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point()
```
Let's break down how that code maps onto the three essential pieces of a ggplot graph listed above.

1. We use `ggplot()` to let R know we're making a ggplot graph (note: the *package* name is `ggplot2`, but the function we use is just `ggplot()`).
2. Within the `ggplot()` function, we have two arguments: The first, `data = penguins`, specifies which object in our environment (in this case, a data frame called penguins) we will get data from for our plot. The second argument, `aes(x = bill_length_mm, y = bill_depth_mm)`, tells us which variables map on to the x- and y-axis aesthetics (hence the `aes()`, for aesthetics).
3. We then add a layer (note the `+` symbol*) containing the specific geometry, or type of graph we're trying to make. A standard scatterplot is made with `geom_point()`.

\*Note: A common mistake is using the pipe operator (`%>%`) between ggplot layers instead of a `+` sign. Try to remember that we build a ggplot graph by adding layers to it piece by piece (`+`), not feeding graph pieces *into* a subsequent layer (which the `%>%` implies).

Let's make a few more bare minimum ggplots. For each, ask yourself: how does this example code map onto the three main pieces of a basic ggplot graph listed above? 

**Example:** a jitterplot (`geom_jitter()`) of flipper length by penguin species. 

```{r}
ggplot(data = penguins, aes(x = species, y = flipper_length_mm)) +
  geom_jitter()
```

**Example:** A boxplot of penguin body mass by island.

Let's add some color, too, just for the halibut. Generally:

- Update line and point colors with `color = `
- Update shape fill colors with `fill = `

```{r}
ggplot(data = penguins, aes(x = body_mass_g, y = island)) +
  geom_boxplot(color = "purple", fill = "cyan2")
```

**Critical thinking:** What might a person unfamiliar with this data conclude if they only saw the boxplot above? Why should this concern us? What critical information is missing that is required for responsible consideration? See also: correlation is not causation, omitted variable bias.

Not all ggplot graph types require two variables. For example, a histogram only wants one numeric variable - it will find the frequencies (plotted on the y-axis) for you. For example, code to make a histogram of penguin flipper lengths is: 

```{r}
ggplot(data = penguins, aes(x = flipper_length_mm)) +
  geom_histogram()
```


### `seaborn` bare minimum

NOTE TO SELF: I used reticulate::py_install("seaborn") while in my project to get seaborn in the right place...

```{python}
import seaborn as sns
```

## Mapping variables onto graph aesthetics

In the examples above, we specified the x- and/or y- axis variables, but didn't update any other aesthetics that would indicate groupings or values within the data. For example, we may want to:

- Differ point color by penguin *species*
- Increase point size by penguin *bill depth*
- Change the boxplot fill color by *island*

When we want to change an aesthetic based on variable values (remember: when I say "values" here, that can mean characters like penguin species), we are *mapping a variable onto a graph aesthetic*. In other words, we are changing an aesthetic of the graph to depend on values within a variable.

### Mapped variable aesthetics in `ggplot2`

We map additional variables onto graph aesthetics with additional `aes()`. 

**Important: Any time you update a graph aesthetic based on a variable value, it will be within `aes()`. Any time you update a graph aesthetic based on a constant (e.g. `color = "blue"`), it should NOT be within `aes()`.**

**Example:** Create a scatterplot of penguin flipper length versus body mass, with color indicating penguin *species* and with point size changing based on the body mass variable value (using the `size` element). 

```{r}
ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, size = body_mass_g))
```
**Example:** Create a violin plot of body mass by species, with the violin fill color dependent on species *and* the violin outline color set to "red".

```{r}
ggplot(data = penguins, aes(x = species, y = body_mass_g)) +
  geom_violin(aes(fill = species), color = "red")
```



## Essential customization

The section above introduces the bare minimum elements needed to create a graph. However, they are far from complete until we update some essential graph elements for clarity. 

Those include: 

- Update axis labels (include label and units)
- If necessary, add a title
- Change the theme

